% Document/LinearAlgebra/VectorSpaces/Foundations/OrderedBases.tex
% Section: Ordered Bases and Coordinates

\newpage
\section{Ordered Bases and Coordinates}

\begin{remark}[Motivation]
    A basis $B$ determines unique coefficients for each vector, but without an ordering
    on $B$, we cannot write these coefficients as a \emph{column vector}. Writing is inherently
    order-preserving, so to represent vectors in matrix form, we need \textbf{ordered bases}.
\end{remark}

\begin{definition}[Ordered Basis]\label{def:ordered-basis}
    An \textbf{ordered basis} of $V$ is a basis $B$ equipped with a \textbf{well-ordering} $<$ on it.
    When $B$ is finite, we write $\mathcal{B} = (b_1, \ldots, b_n)$ as an $n$-tuple.
    When $B$ is infinite, we write $\mathcal{B} = (b_i)_{i \in I}$ for some well-ordered index set $I$.
\end{definition}

\begin{remark}[Ordering Matters]
    The same underlying set with different orderings gives \emph{different} ordered bases.
    For example, $\{e_0, e_1\}$ in $\bb{R}^2$ yields two ordered bases: $(e_0, e_1)$ and $(e_1, e_0)$.
    These give different coordinate representations for the same vector.
\end{remark}

\begin{definition}[Coordinates]\label{def:coordinates}
    Let $\mathcal{B} = (b_i)_{i \in I}$ be an ordered basis for $V$ (with $I$ well-ordered).
    For $v \in V$, the unique representation
    \[
        v = \sum_{i \in F} a_i b_i \quad (\text{$F \subseteq I$ finite, $a_i \neq 0$})
    \]
    determines the \textbf{coordinate vector} $\CoordVec[\mathcal{B}]{v} \in K^{(I)}$ defined by
    \[
        \CoordVec[\mathcal{B}]{v}(i) =
        \begin{cases}
            a_i & \text{if } i \in F \\
            0 & \text{otherwise}
        \end{cases}
    \]
    When $I$ is finite, we write $\CoordVec[\mathcal{B}]{v} = \ColVec{a_0 \\ \vdots \\ a_{n-1}}[\mathcal{B}] \in K^n$.
\end{definition}

\begin{example}[Coordinate Dependence on Ordering]\label{ex:coord-ordering}
    In $\bb{R}^2$, let $v = (3, 2)$. With the standard basis:
    \begin{itemize}
        \item $\mathcal{B}_1 = (e_0, e_1)$: $\CoordVec[\mathcal{B}_1]{v} = \ColVec{3 \\ 2}[\mathcal{B}_1]$
        \item $\mathcal{B}_2 = (e_1, e_0)$: $\CoordVec[\mathcal{B}_2]{v} = \ColVec{2 \\ 3}[\mathcal{B}_2]$
    \end{itemize}
    The coordinates depend on the \textbf{ordering}, not just the set of basis vectors.
\end{example}

\begin{theorem}[Coordinate Bijection]\label{thm:coord-bijection}
    Let $\mathcal{B} = (b_i)_{i \in I}$ be an ordered basis for $V$. The coordinate map
    \[
        \phi_{\mathcal{B}}: V \to K^{(I)}, \quad v \mapsto \CoordVec[\mathcal{B}]{v}
    \]
    is a bijection. (It is in fact a linear isomorphism; see Theorem~\ref{thm:coord-iso}.)
\end{theorem}

\begin{proof}
    \textbf{Injectivity}: If $\CoordVec[\mathcal{B}]{v} = 0$, then all coefficients are zero, so $v = 0$.

    \textbf{Surjectivity}: For any $f \in K^{(I)}$ with finite support $F$, the vector
    $v = \sum_{i \in F} f(i) b_i$ satisfies $\CoordVec[\mathcal{B}]{v} = f$.
\end{proof}

\begin{proposition}[Canonical Basis of $K^{(I)}$]\label{prop:canonical-basis}
    For any set $I$, define $e_i \in K^{(I)}$ by $e_i(j) = \delta_{ij}$ (Kronecker delta).
    Then $(e_i)_{i \in I}$ is a basis for $K^{(I)}$, called the \textbf{canonical basis}.
\end{proposition}

\begin{proof}
    \textbf{Generating}: For any $f \in K^{(I)}$ with finite support $F$:
    \[
        f = \sum_{i \in F} f(i) \cdot e_i
    \]
    (Verify: both sides agree at every $j \in I$.)

    \textbf{Linear independence}: Suppose $0 = \sum_{i \in F} a_i e_i$ for some finite $F$ with $a_i \neq 0$.
    Evaluating at any $j \in F$: $0 = a_j$, contradiction.
\end{proof}

\begin{remark}[Canonicity is Special to $K^{(I)}$]\label{rem:canonical-basis-special}
    The basis $(e_i)_{i \in I}$ is \textbf{canonical} because it is defined intrinsically by
    the function-space structure of $K^{(I)}$ itself --- no arbitrary choice is involved.
    \textbf{An abstract $K$-vector space $V$ has no canonical basis}: any basis of $V$ requires a
    choice (equivalently, a choice of isomorphism $V \cong K^{(I)}$). Even $\bb{R}^n$ only has
    a ``standard'' basis because we keep track of the identification $\bb{R}^n = K^{(n)}$ where $n = \{0, 1, \ldots, n-1\}$.
    The canonical basis exists \emph{only} for spaces of the form $K^{(I)}$.
\end{remark}

\begin{corollary}[Dimension of $K^{(I)}$]\label{cor:dim-finite-support}
    $\Dim[K]{K^{(I)}} = \Card{I}$.
\end{corollary}

\begin{example}[Canonical Bases]\label{ex:canonical-bases}
    Spaces of the form $K^{(I)}$ have a canonical basis. The following are instances:
    \begin{itemize}
        \item $K^n = K^{(n)}$ where $n = \{0, 1, \ldots, n-1\}$: canonical basis $(e_0, e_1, \ldots, e_{n-1})$
        \item $K^{(\kappa)}$ for a cardinal $\kappa$: canonical basis $(e_\alpha)_{\alpha < \kappa}$
    \end{itemize}
    Other vector spaces have \emph{no} canonical basis. For instance, $K[X]$ has a natural ordered
    basis $(1, X, X^2, \ldots)$, but this is only ``natural'' because we implicitly choose the
    isomorphism $K[X] \cong K^{(\bb{N})}$ sending $X^n \mapsto e_n$. A different isomorphism would
    yield a different basis. The basis is not canonical --- only the isomorphism type is.
\end{example}

\begin{remark}[Cardinal Properties of $K^{(I)}$]
    We study the cardinal properties of $K^{(I)}$:
    \begin{enumerate}
        \item \textbf{Lower bound}: Embeddings $k \mapsto k \cdot e_{i_0}$ and $i \mapsto e_i$ give
              $\max(\Card{K}, \Card{I}) \leq \Card{K^{(I)}}$.
        \item \textbf{Bijection}: $K^{(I)} \sim \BigUnion[n \in \bb{N}] (K \setminus \{0\})^n \times \binom{I}{n}$ (disjoint union;
              a vector is determined by its finite support and nonzero values).
        \item \textbf{Finite case}: When both $K$ and $I$ are finite, $K^{(I)} = K^I$, so $\Card{K^{(I)}} = \Card{K}^{\Card{I}}$.
        \item \textbf{Infinite case}: When $\kappa = \max(\Card{K}, \Card{I}) \geq \aleph_0$:
              \[
                  \Card{K^{(I)}} = \sum_{n \in \bb{N}} \kappa^n = \aleph_0 \cdot \kappa = \kappa
              \]
    \end{enumerate}
\end{remark}
